---
title: "Latent factor GWAS and fine-mapping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example_Part2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Here we provide an annotated script for 

1. Estimation of latent factor GWAS summary statistics from observed trait GWAS summary statistics of NMR metabolites from the INTERVAL study, but it is broadly applicable to other data sets. This includes the crucial step of harmonising GWAS to have the same variants present in all GWAS and for all GWAS to have the same effect alleles at each variant. 

2. Identification of latent factor signals and linked observed traits with signals, followed by fine-mapping

This script assumes that the user has created a file "region_list.tsv" containing a list of regions with columns having names "CHR", "START", "END", and each row gives the region's chromosome and start and end positions in base-pair units.
For example, in our NMR analysis, we created 1MB regions that were centered around lead SNPs from the observed trait GWAS.

# Required R libraries and input
```{r,eval=FALSE} 
library(bigsnpr)
library(vroom)
library(propagate)
library(Rsamtools)
library(tidyverse)
library(flashfmZero)


 # directory to save temporary intermediate files - can delete later
OUTPUT_DIR="output/"
save.path=sprintf("%stmpDIR",OUTPUT_DIR)


## Required input files:

# Trait_cov.csv: trait covariance matrix for the observed traits, with traits listed in row and column names
covY <- read.csv("Trait_cov.csv",row.names=1) # trait covariance matrix 

# region_list.tsv: regions for fine-mapping, with columns having names "CHR", "START", "END", and each row gives the region's chromosome and start and end positions in base-pair units
region_list<-read_tsv("region_list.tsv",col_names=TRUE)
regions <- GRanges(seqnames = region_list$CHR , ranges = IRanges(start = region_list$START, end = region_list$END))
chromosome<-as.character(seqnames(regions[array_index]))

# Factor_loading_matrix.csv: factor loading matrix 
FAloadings <- read.csv("Factor_loading_matrix.csv",row.names=1) # factor loadings with factor names in columns and observed traits in row names

# Re-scaled-Factor_loading_matrix.csv: re-scaled factor laoding matrix, i.e contributions matrix with factor names in columns and observed traits in row names
rescaledFAloadings <- read.csv("Re-scaled-Factor_loading_matrix.csv",row.names=1) 

# observed trait GWAS summary statistics
SUM_STAT_DIR="results/metabolomics/nmr/raw_results_tabixd/nmr"


# LD matrix for region - here bgen files are used
GENO_DIR="data/interval_imputed_bgen_layout2"
bgenfiles = sprintf("%s/impute_%s_interval_l2_8bit.bgen",GENO_DIR, chromosome)
backingfile = sprintf("%s/lipid/bigSNP/%d",OUTPUT_DIR,array_index)

# sample size of study
N <- 40849


# directory to save fine-mapping results
file_FMresults <- sprintf("output/%d.Rdata",array_index) # e.g. output file "1.Rdata" will contain results for region 1 (row 1 of region_list.tsv)

```


# Harmonise GWAS data from observed traits

```{r,eval=FALSE} 
# Specify which row of the "region_list.tsv" file to select region for analysis 
array_index <- as.integer(Sys.getenv("SLURM_ARRAY_TASK_ID")) # region number as provided by job array submission on hpc

phenotype_list <- rownames(covY) # trait names
P<-length(phenotype_list) # number of traits

# Read in GWAS summary statistics 
obsgwas <- vector("list", P)
for(i in 1:P) {
	print(i)
	phenotype<-phenotype_list[i]
	tabix<-TabixFile(sprintf("%s/%s/%s_chr%s_imputed.gz", SUM_STAT_DIR, phenotype, phenotype, chromosome))
	result<-scanTabix(tabix, param = regions[array_index])[[1]]
	result_tibble <- as_tibble(do.call(rbind, stringr::str_split(result, "\t")),  .name_repair = "minimal")
	colnames(result_tibble)<-c("SNP","CHR","BP","GENPOS","ALLELE1","ALLELE0","A1FREQ","INFO","BETA","SE","P_BOLT_LMM_INF","P_BOLT_LMM")
	result_tibble$rsID<-sprintf("%s_%s", result_tibble$CHR, result_tibble$BP) # no alleles in snp ids in case need to flip
	obsgwas[[i]]<-as.data.frame(result_tibble)
}
names(obsgwas) <- phenotype_list

# Harmonise so all traits have the same variants and effect alleles, and remove duplicates 
# In this example, the GWAS files are in the BOLT-LMM format, but the code is flexible to other column names
# Specify your column that contains your effect estimates in the beta_colname argument, effect allele frequency column name in EAfreq_colname, p-value column name in pvalue_colname, effect allele column name in EA_colname, and non-effect allele column name in NEA_colname.
# Specify the minimum MAF and minimum INFO score in minMAF and minINFO. Defaults are minMAF=0.005,minINFO=0.4.
# The output GWAS will have column names "beta", "SE", "rsID", and "EAF", for the effect estimates and their standard errors, variant IDs, and effect allele frequencies. These names are needed to fun the fine-mapping functions.
ooo <- harmoniseGWAS(obsgwas,minMAF=0.005,minINFO=0.4,beta_colname="BETA",EAfreq_colname="A1FREQ",pvalue_colname="P_BOLT_LMM_INF",EA_colname="ALLELE1",NEA_colname="ALLELE0")
obsgwas <- ooo
rm(ooo) 
print("GWAS reading in and formatting done")

```

# Estimate latent factor GWAS from summary statistics
```{r,eval=FALSE} 
# first ensure that covY matrix and observed trait GWAS have same ordering as latent factor loadings and contributions
covY <- covY[rownames(FAloadings),rownames(FAloadings)]
obsGWAS <- obsgwas[rownames(FAloadings)]


# latent gwas
latent_gwas = latentGWAS(obsGWAS = obsGWAS,covY = covY,L = FAloadings, beta_colname="beta",se_colname="SE",
                       snpID_colname="rsID",EAF = obsGWAS[[1]]$EAF)
print("latent GWAS done")
```

# Identify latent factors and their linked traits that have a signal

```{r,eval=FALSE} 
# identify latent traits with a GWS signal in region
check <- sapply(latent_gwas,function(x) any(x$p_value<5E-8))
fm_traits_latent <- names(check[check==TRUE]) # latent traits with a GWS signal in region

print("fm_traits_latent")
print(fm_traits_latent)

# identify traits with contribution >0.20 from latent traits to be fine-mapped and that have a GWAS signal in region

if(length(fm_traits_latent)==1) {
	tmp <- rescaledFAloadings[,fm_traits_latent] 
	linked_traits <- rownames(rescaledFAloadings)[which(tmp>0.2)] # identify traits with contribution >0.20 from latent traits to be fine-mapped
	check <- sapply(obsGWAS[linked_traits],function(x) any(x$p_value<5E-8))
	fm_traits_obs <- names(check[check==TRUE]) # obs traits with a GWS signal in region and contrib > 0.2 with latent factors fine-mapped
	print("fm_traits_obs")
	print(fm_traits_obs)

}

if(length(fm_traits_latent)>1) {
	tmp <- rescaledFAloadings[,fm_traits_latent] 
	linked_traits <- rownames(tmp)[which(apply(tmp,1, function(x) any(x>0.2)))] # identify traits with contribution >0.20 from latent traits to be fine-mapped
	check <- sapply(obsGWAS[linked_traits],function(x) any(x$p_value<5E-8))
	fm_traits_obs <- names(check[check==TRUE]) # obs traits with a GWS signal in region and contrib > 0.2 with latent factors fine-mapped
	print("fm_traits_obs")
	print(fm_traits_obs)
}

# If no latent factor has a signal or no linked observed trait has a signal, then stop running script
# because nothing to fine-map.
try(if(length(fm_traits_latent)==0) stop("No latent factor has a signal"))
try(if(length(fm_traits_obs)==0) stop("No observed trait has a signal"))
```

# Read in genotype data for LD - needed for fine-mapping

This step assumes that bgen files for LD calculation are available, either from in-sample data our a genetically similar reference panel. See below this block of code for an alternative if LD is pre-calculated from a different software.

```{r,eval=FALSE} 
bgenfiles = sprintf("%s/impute_%s_interval_l2_8bit.bgen",GENO_DIR, chromosome)
backingfile = sprintf("%s/lipid/bigSNP/%d",OUTPUT_DIR,array_index)
list_snp_id=list() # Each SNP ID should be in the form "<chr>_<pos>_<a1>_<a2>"
list_snp_id[[1]] = paste0(obsGWAS[[1]]$CHR,'_',
                          obsGWAS[[1]]$BP,'_',
                          obsGWAS[[1]]$EA,'_',
                          obsGWAS[[1]]$NEA)


rds = snp_readBGEN(
  bgenfiles,
  backingfile,
  list_snp_id,
  ind_row = NULL,
  bgi_dir = dirname(bgenfiles),
  read_as = "dosage",
  ncores = 1
)

test <- snp_attach(rds)

print("map")
print(head(test$map))

RPinfo = test$map[,c(1,2,4,5,6)]
RPinfo = as.data.frame(RPinfo)
names(RPinfo)[2:3] = c("rsID","BP")  # c("chromosome","rsID","BP","allele1","allele2") 
RPinfo$rsID <- paste0(chromosome,"_",RPinfo$BP) # no alleles in snp ids in case need to flip

genotypes = test$genotypes[1:dim(test$genotypes)[1],1:dim(test$genotypes)[2]]
colnames(genotypes) = RPinfo$rsID

```

If LD is pre-calculated from other software, just read in your LD matrix file and a file that contains the columns: "chromosome","rsID","BP","allele1","allele2" coinciding with your LD data source. Here, it is assumed that LD_1.csv is the LD matrix for region 1.

```{r,eval=FALSE} 
corX <- read.csv(sprintf("LD_%s.csv", array_index),row.names=1) # use row.names=1 if your file has row names in column 1 
# file that contains the columns: "chromosome","rsID","BP","allele1","allele2" coinciding with your LD data source
# It MUST have SNP rows in the same order as your LD file.
RPinfo <- read.csv(sprintf("LDinfo_%s.csv", array_index))
RPinfo$rsID <- paste0(chromosome,"_",RPinfo$BP) # to match with snp ids used in gwas
rownames(corX) <- colnames(corX) <- RPinfo$rsID
```

# Align GWAS to LD

This step is needed to ensure that the GWAS and LD source have the same effect alleles or consistently have opposite effect alleles. Otherwise, the correlation direction will sometimes be incorrect, which will affect dine-mapping results.

```{r,eval=FALSE} 
# keep only traits that will be fine-mapped
tmp1 <- latent_gwas[fm_traits_latent] 
tmp2 <- obsGWAS[fm_traits_obs]

# align obsGWAS and latent_gwas to alleles coded for LD
sig_gwas_latent <- vector("list",length(fm_traits_latent))
for(i in 1:length(fm_traits_latent)) {
 print(i)
 print(head(tmp1[[i]]))
 sig_gwas_latent[[i]] <- alignGWAS(tmp1[[i]],RPinfo)
}

sig_gwas_obs <- vector("list",length(fm_traits_obs))
for(i in 1:length(fm_traits_obs)) {
 print(i)
 print(head(tmp2[[i]]))
 sig_gwas_obs[[i]] <- alignGWAS(tmp2[[i]],RPinfo)
}
print("align LD and GWAS done")


```

# LD calculation 

Only needed if calculating LD here.

```{r,eval=FALSE} 
# snps present in all gwas and in LD
isnp <- intersect(sig_gwas_obs[[1]]$rsID, colnames(genotypes) )
print(c("isnp",length(isnp)))

X <- genotypes[,isnp]
for(i in 1:length(fm_traits_obs)) {
 rownames(sig_gwas_obs[[i]]) <- sig_gwas_obs[[i]]$rsID
 sig_gwas_obs[[i]] <- sig_gwas_obs[[i]][isnp,]
}
for(i in 1:length(fm_traits_latent)) {
  rownames(sig_gwas_latent[[i]]) <- sig_gwas_latent[[i]]$rsID
  sig_gwas_latent[[i]] <- sig_gwas_latent[[i]][isnp,] 
}



Xqc <- apply(X,2,LDqc,theta=0.2,BestGuess=TRUE) # best-guess genotype matrix, using dosage tolerance 0.2
# i.e.only dosages d that satisfy d<0.2, 0.8<d<1.2, and d>1.8 are retained; otherwise set to NA

Nprop <- apply(Xqc,2,function(x) sum(!is.na(x)))/nrow(Xqc) # identify and keep variants that have genotype within tolerance for at least 80% of the sample
keep <- which(Nprop >= 0.80)

print(length(keep))

Xqc <- Xqc[,keep]
for(i in 1:length(sig_gwas_obs)) sig_gwas_obs[[i]] <- sig_gwas_obs[[i]][keep,]
for(i in 1:length(sig_gwas_latent)) sig_gwas_latent [[i]] <- sig_gwas_latent[[i]][keep,] 

ldout <- bigcor(Xqc, size=min(2000, ncol(Xqc)), verbose=FALSE, fun= "cor", use="p" )
corX <- as.matrix(ldout[1:dim(ldout)[1],1:dim(ldout)[1]])
rownames(corX) <- colnames(corX) <- colnames(Xqc)

```


# Data formatting for fine-mapping

```{r,eval=FALSE} 
# Set same snps in LD and gwas
isnp <- intersect(rownames(corX), sig_gwas_obs[[1]]$rsID)
for(i in 1:length(sig_gwas_obs)) sig_gwas_obs[[i]] <- sig_gwas_obs[[i]][isnp,]
for(i in 1:length(sig_gwas_latent)) sig_gwas_latent [[i]] <- sig_gwas_latent[[i]][isnp,] 

# change snp names so that they start with a characater - this is a requirement for JAM
corXnames <- paste0("chr",rownames(corX))
rownames(corX) <- colnames(corX) <- corXnames

sig_gwas_latent_names <- paste0("chr",sig_gwas_latent[[1]]$rsID)
for(i in 1:length(sig_gwas_latent)) sig_gwas_latent[[i]]$rsID <- sig_gwas_latent_names

sig_gwas_obs_names <- paste0("chr",sig_gwas_obs[[1]]$rsID)
for(i in 1:length(sig_gwas_obs)) sig_gwas_obs[[i]]$rsID <- sig_gwas_obs_names

names(sig_gwas_obs) <- fm_traits_obs
names(sig_gwas_latent) <- fm_traits_latent



```

# Fine-mapping

**NOTE 1**: If running on the hpc, either

1. Set argument NCORES=length(fm_traits_latent) in FLASHFMZEROwithJAMd for parallelisation, and set NCORES=length(fm_traits_obs) in multiJAMd. Could also set a smaller number for NCORES, but setting to the number of traits is optimal.
OR
2. Set NCORES=1 in FLASHFMZEROwithJAMd and multiJAMd.


**NOTE 2**: In all fine-mapping functions below, set jam.nM.iter =1 if you have in-sample LD. For out-of-sample LD, like from 1000 Genomes, set jam.nM.iter =5.


```{r,eval=FALSE} 
# If one latent factor, then run JAMdwithGroups; if multiple latent factors, then run FLASHFMZEROwithJAMd, which will output both single and multi-trait results
if(length(fm_traits_latent)==1){
  FM_latent <- JAMdwithGroups(sig_gwas_latent[[1]], N=N, corX, save.path, cred = 0.99,
                                                   jam.nM.iter =1, maxcv = 1, maxcv_stop = 20, 
                                                   min.mppi = 0.01, r2.minmerge = 0.8)
 FM_latent_CS99 <- FM_latent$CS
} else {
	FM_latent <- FLASHFMZEROwithJAMd(sig_gwas_latent, 
                                                     corX, 
                                                     N = N, 
                                                     save.path=save.path, 
                                                     TOdds = 1,
                                                     cpp = 0.99, 
                                                     NCORES=1, 
                                                     maxcv = 1, 
                                                     maxcv_stop = 20, 
                                                     jam.nM.iter = 1, 
                                                     r2.minmerge = 0.8, 
                                                     minsnpmppi = 0.01)
                                                 
  FM_latent_CS99 <- allcredsetsPP(FM_latent$mpp.pp,cred=.99) # construct 99% credible sets (change cred=0.95 for 95% credible set)
    names(FM_latent_CS99$fm) <- fm_traits_latent
  names(FM_latent_CS99$flashfm) <- fm_traits_latent
}


 #  for observed trait(s) - if one trait, then run JAMdwithGroups; if multiple traits, then run multiJAMd
if(length(fm_traits_obs)==1){
  FM_obs <- JAMdwithGroups(sig_gwas_obs[[1]], N=N, corX, save.path, cred = 0.99,
                                                    jam.nM.iter =1, maxcv = 1, maxcv_stop = 20, 
                                                    min.mppi = 0.01, r2.minmerge = 0.8)
  FM_obs_CS99 <- FM_obs$CS 

} else {
 FM_obs <- multiJAMd(sig_gwas_obs,  corX, N=N, save.path,
                                    maxcv = 1, maxcv_stop = 20, jam.nM.iter =1, r2.minmerge=0.8, minsnpmppi = 0.01,
                                    NCORES=1) #if running on the hpc
                                   
   FM_obs_CS99  <- multiJAMdCS(FM_obs, cred = 0.99)
   names(FM_obs_CS99$fm) <- fm_traits_obs
}
 
# save: observed trait names,latent factor names, latent factor fine-mapping results, observed trait fine-mapping results, observed trait credible sets, latent factor credible sets, 
save(fm_traits_obs, fm_traits_latent, FM_latent, FM_obs, FM_obs_CS99, FM_latent_CS99, file=file_FMresults)

```

# Construct summary table of fine-mapping results

This is an example of how to use built-in functions from this package to summarise fine-mapping results from many traits and regions, producing a table as in Supplementary Tables S9 and S14 of our associated paper. "FMsummary_table_general" is a wrapper function for "FMsummary_table". 

The below code will loop through all regions in the regions file "region_list.tsv" and loads in the corresponding fine-mapping output from the file "array_index".Rdata, e.g. 1.Rdata is the file output for region 1. It assumes that the output is saved with names as above, i.e. fm_traits_obs, fm_traits_latent, FM_latent, FM_obs, and will caclulate the 99/% credible sets for each trait/factor (for 95/% credible sets use "cred=0.95" in FMsummary_table_general)

The code below checks if the file "array_index".Rdata exists. This pipeline is designed such that the script will return an error and no output if there are not latent factor AND observed trait signals in the region, so it is possible for "array_index".Rdata not to exist.

```{r,eval=FALSE} 
out <- NULL
regions <- read.table("region_list.tsv",header=TRUE)
for(array_index in 1:nrow(regions)) {
  fname <- paste0(array_index,".Rdata") # adjust this file name if needed
  if(file.exists(fname)){
  load(fname)
  out1 <- FMsummary_table_general(FM_obs, FM_latent, fm_traits_ob, fm_traits_latent,regions=regions,array_index=array_index,cred=0.99) 
  out <- rbind(out,out1)
  }
}

```


